\chapter{Publications and Presentations}

\section*{List of First Author Publications}

The following papers are included in full as part of, or in lieu of, chapters within this dissertation:

\begin{itemize}

\item \textbf{A Framework for Dialogue-Based Human-AI Creative Collaboration.} Ocampo, Rodolfo, Bown, Oliver, \& Grace, Kazjon. (2022). In \textit{Proceedings of the CHI 2022 Workshop on Generative AI and HCI}. New Orleans, USA. (paper)

\item \textbf{Beyond Chat: Towards Greater User Involvement and Agency in Human-AI Co-Creativity through Shared Collaborative Spaces.} Ocampo, Rodolfo, Bown, Oliver, \& Grace, Kazjon. (in press). Accepted for publication in \textit{Proceedings of the International Conference on Computational Creativity (ICCC) 2025}, Campinhas, Brazil, June 23–27.

\item \textbf{Integrating Generative AI into Creative Workflows: Dealing with Consistency, Scene Control, and Refinement in a Professional Image Generation Case Study.} Ocampo Blanco, Rodolfo, \& Bown, Oliver. (2024). In \textit{Proceedings of the International Conference on Computational Creativity (ICCC)}. Jönköping, Sweden. (paper)

\item \textbf{Using GPT-3 to Achieve Semantically Relevant Data Sonification for an Art Installation.} Ocampo, Rodolfo, Bown, Oliver, Wright, Brendan, Shave, Justin, \& Pegram, Caroline. (2023). In \textit{Artificial Intelligence in Music, Sound, Art and Design: 12th International Conference}. Brno, Czech Republic. (paper)

\item \textbf{Interpretative Data Sonification: Using LLMs to Interpret Data and Generate Continuous Soundscapes at the Sydney Opera House.} Ocampo, Rodolfo, Bown, Oliver, Wright, Brendan, Shave, Justin, \& Pegram, Caroline. (2024). In \textit{Proceedings of the 21st Sound and Music Computing Conference (SMC 2024)}. Porto, Portugal: ESMAE. ISBN: 978-989-35207-5-8.


\end{itemize}



\section*{List of Co-Author Publication}

The following publications are referenced throughout this thesis. 

\begin{itemize}

\item \textbf{The Human–Built Environment–Natural Environment Relation: An Immersive Multisensory Exploration with “System of a Sound.”} Andres, Josh, Ocampo, Rodolfo, Bown, Oliver, Hill, Charlton, Pegram, Caroline, Schmidt, Adrian, Shave, Justin, \& Wright, Brendan. (2023). In \textit{Proceedings of the 28th International Conference on Intelligent User Interfaces (IUI 2023)}, March 27. ACM, New York, NY, USA. \url{https://doi.org/10.1145/3581754.3584119}


\item \textbf{“Coffee so good, you’ll want to slap your barista”: Evaluating AI Co-Creation through Dialogic Actions.} Bown, Oliver, Grace, Kazjon, Ocampo Blanco, Rodolfo, \& Ibarrola, Francisco. (2025). \textit{Creativity Research Journal}, June, pp. 1–17. DOI: \href{https://doi.org/10.1080/10400419.2025.2501735}{10.1080/10400419.2025.2501735}. URL: \url{https://doi.org/10.1080/10400419.2025.2501735}


\end{itemize}

\section*{List of Presentations}

\subsection*{Oral presentations:} 

\begin{itemize}
\item Evolutionary Art and Music Conference (Evo*), Brno, Czech Republic, 2023
\item Sound and Music Computing Conference, Porto, Portugal, 2024
\item ACM CHI Conference on Human Factors in Computing Systems, Japan, 2025


\end{itemize}


\subsection*{Performances and installations}

\begin{itemize}
    \item "Music of the Sails", Sydney Opera House, 2023
    \item "System of a Sound",Cybernetic Serendipty, Australian National University, 2022
\end{itemize}

\subsection*{Poster presentations:}

\begin{itemize}
    \item International Conference of Computational Creativity (ICCC), Jonkoping, Sweden 2024
\end{itemize}


